---
title: "DATA 606 Data Project Proposal"
author: "Tyler Frankenberg"
date: "04/11/2021"
output: pdf_document
---

```{r load_packages}

library(tidyverse)
library(jsonlite)

```

### Data Preparation

```{r setup, echo=TRUE, results='hide', warning=FALSE, message=FALSE}

# read in data from JSON file
url <- "https://raw.githubusercontent.com/curdferguson/606-Project/main/winemag-data-130k-v2.json"
wines <- fromJSON(url)

# extract vintage year from the title column for wines with years after 1900
wines <- wines %>% mutate(
  vintage = str_extract(wines$title, "(19|20)[0-9]{2}")
)

# drop unneeded columns
wines <- wines %>% select(!c(description, taster_name, taster_twitter_handle, designation, region_2))

# drop the rows of wines without vintages or with unreliable vintage years (pre-1900)
wines <- wines %>% filter(is.na(vintage) == FALSE & is.na(price) == FALSE & is.na(points) == FALSE)

#convert vintage to an ordered factor and points to numeric
wines$vintage <- ordered(wines$vintage)
wines$points <- as.numeric(wines$points)

head(wines)

```


### Research question 

**You should phrase your research question in a way that matches up with the scope of inference your dataset allows for.**

Does a higher price predict a higher *Wine Enthusiast* rating?  Is the vintage (year) predictive of a higher *Wine  Enthusiast* rating?

### Cases 

**What are the cases, and how many are there?**

There are 125345 cases.  Each case represents a review in *Wine Enthusiast*, published in November 2017.  Each review concerns a single wine.


### Data collection 

**Describe the method of data collection.**

The dataset was scraped from *Wine Enthusiast* by Kaggle user `zackthoutt` <https://www.kaggle.com/zynicide> in November 2017.


### Type of study 

**What type of study is this (observational/experiment)?**

This is an observational study.


### Data Source 

**If you collected the data, state self-collected. If not, provide a citation/link.**

The dataset is obtained from Kaggle through link: <https://www.kaggle.com/zynicide/wine-reviews>. It is published under a Creative Commons BY-NC-SA 4.0 license <https://creativecommons.org/licenses/by-nc-sa/4.0/>.


### Dependent Variable

**What is the response variable? Is it quantitative or qualitative?**

The response variable, `points`, is quantitative.


### Independent Variable

**You should have two independent variables, one quantitative and one qualitative.**

The first independent variable, `price`, is quantitative, and the second independent variable `vintage` is qualitative.


### Relevant summary statistics 

**Provide summary statistics for each the variables. Also include appropriate visualizations related to your research question (e.g. scatter plot, boxplots, etc). This step requires the use of R, hence a code chunk is provided below. Insert more code chunks as needed.**

```{r}

# construct and display points summary
points_summary <- wines %>% summarize(
  mean_points = mean(points),
  sd_points = sd(points),
  min_points = min(points),
  max_points = max(points)
)

points_summary

# show histogram of price distribution
wines %>% ggplot(aes(x = points)) +
  geom_histogram(bins = 30) +
  theme(legend.position = "none") +
  labs(title= "Distribution of Review Points", 
       subtitle= "Response Variable", 
       x = "Points", 
       y ="# of Reviews", 
       caption="")

# construct and display price summary
price_summary <- wines %>% summarize(
  mean_price = mean(price),
  sd_price = sd(price),
  min_price = min(price),
  max_price = max(price)
)

price_summary

# show histogram of price distribution
wines %>% ggplot(aes(x = price)) +
  geom_histogram(bins = 30) +
  scale_x_log10() + 
  theme(legend.position = "none") +
  labs(title= "Distribution of Wine Prices", 
       subtitle= "Explanatory Variable - Quantitative", 
       x = "Price (Scale = log10)", 
       y ="# of Wines", 
       caption="")


# show histogram of vintage distribution
wines %>% ggplot(aes(x = vintage)) +
  geom_bar() +
  scale_x_discrete(breaks=c("1904", "1934", "1957", "1969", "1980", "1990", "2000", "2010")) +
  scale_y_log10() + 
  theme(legend.position = "none") +
  labs(title= "Distribution of Wine Vintages", 
       subtitle= "Explanatory Variable - Qualitative", 
       x = "Vintage", 
       y ="# of Wines (Scale = log10)", 
       caption="")


```

